---
title: "Part 2: Quantitative data analysis"
subtitle: "Quantitative Session of the 2026 HUSOE ELPS Methods and Data Institute"
author: "Nathan Alexander, PhD"
institute: 
  - "School of Education"
  - "Center for Applied Data Science and Analytics"
format: 
  html: default
  revealjs:
    output-file: part02-slides.html
    height: 900
    width: 1600
    smaller: false
    slide-number: c/t #< collapsted/total
    logo: "../img/howard-logo.jpg"
    footer: "[Workshop Website](https://quant-shop.github.io/elps-data-methods-institute/)"
    toc: false
    echo: true
    incremental: false
  pdf:
    toc: true
    number-sections: true
    colorlinks: true
    geometry: 
      - margin=1in
geometry: margin=1.0in
editor_options:
  markdown:
    wrap: sentence
---

```{r setup, include=F, warning=F, message=F}
library(here)
here::i_am("parts/part02.qmd")
```

### The Big Picture

Prior to jumping into more details about data analysis, and how we might be more critical in our anlayses, it will be important to gather a "big picture" view of the field. This big picture view will provide us with a high-level description of the various topics we will cover in the course.

---

#### Population parameter

A number which summarizes the entire group.

#### Sample statistic

A single number that summarizes a subset of data, or the sample.

---

#### Confidence interval (CI)

-   Range of likely or plausible values for a population parameter.

-   Based on a sample and statistics from that sample.

-   *Margin of error* represents the number of standard deviations on a statistic.

---

#### Hypothesis testing

Statistical procedure used to test an existing claim about a population.

-   Test is based on data; most ideally data collected via probability sampling.

-   $H_0$ - Null hypothesis: If data supports the claim, fail to reject $H_0$

-   $H_a$ - Alternative hypothesis: If data does not support claim, reject $H_0$

---

#### Analysis of variance (ANOVA)

Comparing means of more than two populations.

*F-statistic* is a ratio that is used to compare variability between sets.

---

##### Multiple comparison procedures

Set of statistical tests that compare means to each other.

-   Examples include Turkey's test, Least significant difference (LSD), pairwise t-test

-   These tests are only conducted if you analysis of variance identifies differences

---

#### Interaction effects

Interaction effects are relevant to statistical models that use two or more variables.

#### Correlation

Measures the strength and direction of a linear relationship between two variables.

#### Linear regression

Helps make predictions for one variable based on the values of another.

**There are multiple types of regression**:

  - Simple linear regression

  - Multiple linear regression

  - Logistic regression

  - Non-linear regression

---

#### Chi-square test

When using correlation and regression analyses, one core assumption is that the variables are quantitative in nature.

We use a chi-square test to study categorical variables.

---

## Discussion: What types of statistical tests will you be using in your dissertation?

---

## Case Study: HBCUs and Educational Geography

In this section, we analyze data on Historically Black Colleges and Universities (HBCUs) to explore patterns in location, institutional type, and founding history.

------------------------------------------------------------------------

### Step 1: Load Packages

```{r, message=F, warning=F}
#install.packages("tidyverse")
#install.packages("lubridate")
library(tidyverse)
library(lubridate)
```

We'll start with a guided example using data on Historically Black Colleges and Universities.

---

### Step 2: Load the data

```{r, message=F, warning=F}

library(tidyverse)
hbcu <- read_csv("https://raw.githubusercontent.com/quant-shop/intro-comp-educ-soc/refs/heads/main/data/hbcu_data.csv")

# View the first few rows
head(hbcu)

# Check structure
glimpse(hbcu)

# Summary statistics
summary(hbcu)
```

---

### Step 3: Clean the data

```{r, message=F, warning=F}
hbcu <- hbcu %>%
  mutate(
    founded = as.numeric(founded),
    type = as.factor(type),
    state = as.factor(state)
  )
```

---

### Step 4: Basic Exploration

How many institutions are in the data set?

```{r, message=F, warning=F}
nrow(hbcu)
```

What is the distribution by HBCU type?

```{r, message=F, warning=F}
hbcu %>%
  count(type) %>%
  arrange(desc(n))

```

---

### Step 5: Historical Analysis

When were the HBCUs founded?

```{r, message=F, warning=F}
hbcu %>%
  ggplot(aes(x = founded)) +
  geom_histogram(binwidth = 10, fill = "steelblue", color = "white") +
  labs(
    title = "Founding Years of HBCUs",
    x = "Year Founded",
    y = "Number of Institutions"
  )
```

---

### Step 6: Geographic Distribution

HBCUs by state

```{r, message=F, warning=F}
hbcu %>%
  count(state, sort = TRUE)
```

Basic visualization of HBCUs by state

```{r, message=F, warning=F}
hbcu %>%
  count(state) %>%
  ggplot(aes(x = reorder(state, n), y = n)) +
  geom_col(fill = "darkgreen") +
  coord_flip() +
  labs(
    title = "Number of HBCUs by State",
    x = "State",
    y = "Count"
  )
```

---

### Step 7: Mapping

```{r, message=F, warning=F}
#install.packages("maps", repos = "http://cran.us.r-project.org")
library(maps)
```

```{r, message=F, warning=F}
hbcu %>%
  ggplot(aes(x = lon, y = lat)) +
  borders("state") +
  geom_point(color = "red", alpha = 0.6) +
  coord_fixed(1.3) +
  labs(
    title = "Geographic Distribution of HBCUs in the U.S.",
    x = "Longitude",
    y = "Latitude"
  )
```

---

## Case Study: Data on State School Districts

### Step 1: Load the libraries

```{r, message=F, warning=F}
library(dplyr)
library(readr)
library(stringr)
library(janitor)
```

### Step 2: Load the data

```{r, message=F, warning=F}
district <- read_csv(
  "https://raw.githubusercontent.com/quant-shop/intro-comp-educ-soc/refs/heads/main/data/state-district-data.csv",
  show_col_types = FALSE
)

# district <- read_csv("../data/state-district-data.csv")

str(district)
```

### Step 3: Clean the data

```{r, message=F, warning=F}
district_clean <- district %>%
  
  # Make long variable names easier to use
  clean_names() %>%
  
  # Trim whitespace in all text fields
  mutate(across(where(is.character), str_trim)) %>%
  
  # Convert measure columns stored as character text to numeric.
  # Keep state identifiers as character variables.
  mutate(
    across(
      where(is.character) &
        !all_of(c(
          "state_name",
          "state_abbr",
          "state_name_state_2023_24"
        )),
      ~ parse_number(.x, na = c("", "NA", "–", "†"))
    )
  ) %>%
  
  # Keep one state-name variable rather than two identical versions
  select(-state_name_state_2023_24)

glimpse(district_clean)
```

### Step 4: Feature engineering

We can create variables that answer educational questions.

```{r}
str(district_clean)

district_features <- district_clean %>%
  mutate(
    enrollment_per_operational_district =
      if_else(
        total_number_operational_school_districts_district_2023_24 > 0,
        total_enrollment_exclude_ae_for_sy_2014_15_onward_state_2023_24 /
          total_number_operational_school_districts_district_2023_24,
        NA_real_
      ),
    
    schools_per_operational_district =
      if_else(
        total_number_operational_school_districts_district_2023_24 > 0,
        total_number_operational_schools_public_school_2023_24 /
          total_number_operational_school_districts_district_2023_24,
        NA_real_
      ),
    
    charter_school_share =
      if_else(
        total_number_operational_schools_public_school_2023_24 > 0,
        total_number_operational_charter_schools_public_school_2023_24 /
          total_number_operational_schools_public_school_2023_24,
        NA_real_
      ),
    
    elementary_teacher_share =
      if_else(
        full_time_equivalent_fte_teachers_state_2023_24 > 0,
        elementary_teachers_state_2023_24 /
          full_time_equivalent_fte_teachers_state_2023_24,
        NA_real_
      )
  )

district_features %>%
  select(
    state_name,
    total_enrollment_exclude_ae_for_sy_2014_15_onward_state_2023_24,
    pupil_teacher_ratio_state_2023_24,
    enrollment_per_operational_district,
    schools_per_operational_district,
    charter_school_share,
    elementary_teacher_share
  ) %>%
  glimpse()
```

What kinds of questions might you answer with this data set?

## Recommended readings

Knaflic, C. N. (2015). *Storytelling with data: A data visualization guide for business professionals*. Wiley.
