---
title: "DATA 202 - Analyzing Multiple Variables"
author: "FirstName LastName"
output: html_document
editor_options:
    chunk_output_type: console
---

```{r setup, include=F}
knitr:: opts_chunk$set(echo = TRUE)
```

There are multiple options when conducting the analysis of multiple variables (i.e., more than two variables). In this file, I provide some very basic code to help you get started on your analysis for your paper on analyzing the relationship between multiple variables.

# Libraries and packages

Load the libraries we needed. Be sure to `install.packages()` where needed.

```{r, message=F, warning=F}
# again, remember to install necessary packages first
# install.packages("mvnormtest)

# load libraries
library(ggplot2)
library(dplyr)
library(corrplot)
library(car)
library(lmtest)
library(mvnormtest)
library(ggpubr)  # combined plots
library(vcd)     # mosaic plots
```

## Load sample data from github

We'll use the gender pay gap data again that I have prepared.

```{r, message=F, warning=F}
pay_gap_data <- read.csv("https://raw.githubusercontent.com/data-202/sp25/refs/heads/master/data/pay_data.csv")
```

# Exploratory analysis

```{r, message=F, warning=F}
# view the structure of your data
str(pay_gap_data)

# view the top and bottom of your data
head(pay_gap_data)
tail(pay_gap_data)

# gather counts for your categorical data
table(pay_gap_data$gender)
table(pay_gap_data$job_title)

# gather descriptive statistics for your numeric variables
pay_gap_data %>%
  select(years_experience, annual_salary) %>% 
  summary()
```

## Anlayzing bivariate relationships

We then want to analyze the bivariate relationship between each of the variables in our model.

### Numeric-Numeric relationships

We'll start with a scatter plot to visualize the relationship between the numeric variables.

```{r, message=F, warning=F}
pairs(pay_gap_data[, c("annual_salary", "years_experience")],
      main = "Pairwise Scatterplot Matrix")
```

```{r, message=F, warning=F}
ggplot(pay_gap_data, aes(x = years_experience, y = annual_salary)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", color = "red") +
  labs(title = "Experience vs Salary Relationship",
       x = "Years of Experience", y = "Annual Salary")
```

```{r, message=F, warning=F}
# check correlation value
cor.test(pay_gap_data$annual_salary, pay_gap_data$years_experience)
```

### Categorical-Numeric relationships

```{r, message=F, warning=F}
## Gender vs Salary
ggplot(pay_gap_data, aes(x = gender, 
                         y = annual_salary, 
                         fill = gender)) +
  geom_boxplot() +
  labs(title = "Salary Distribution by Gender")
```

```{r, message=F, warning=F}
## Job Title vs Salary (sorted by median salary)
pay_gap_data %>%
  mutate(job_title = reorder(job_title, annual_salary, median)) %>%
  ggplot(aes(x = job_title, y = annual_salary, fill = job_title)) +
  geom_boxplot() +
  coord_flip() +  # Better readability for long job titles
  labs(title = "Salary Distribution by Job Title") +
  theme(legend.position = "none")
```

```{r, message=F, warning=F}
## Job Title vs Experience
pay_gap_data %>%
  mutate(job_title = reorder(job_title, years_experience, median)) %>%
  ggplot(aes(x = job_title, y = years_experience, fill = job_title)) +
  geom_boxplot() +
  coord_flip() +
  labs(title = "Experience Distribution by Job Title") +
  theme(legend.position = "none")
```

### Categorical-Categorical relationship

```{r, message=F, warning=F}
# bar chart of the categorical relationships
ggplot(pay_gap_data, aes(x = job_title, fill = gender)) +
  geom_bar(position = "dodge") +
  coord_flip() +
  labs(title = "Gender Distribution Across Job Titles",
       x = "Job Title", y = "Count") +
  theme_minimal() +
  theme(axis.text.y = element_text(angle = 0, hjust = 1))
```

We can then conduct a chi-square test to see if there is a relation.

```{r, message=F, warning=F}
## Chi-square test
chisq.test(pay_gap_data$gender, pay_gap_data$job_title)
```

## Visualizing the combined relationships

```{r, message=F, warning=F}
## Salary ~ Experience faceted by Gender
ggplot(pay_gap_data, aes(x = years_experience, y = annual_salary)) +
  geom_point(aes(color = gender)) +
  geom_smooth(method = "lm") +
  facet_wrap(~gender) +
  labs(title = "Experience-Salary Relationship by Gender")
```

```{r, message=F, warning=F}
## Salary ~ Experience colored by Job Title
ggplot(pay_gap_data, aes(x = years_experience, y = annual_salary, color = job_title)) +
  geom_point(alpha = 0.7) +
  labs(title = "Experience-Salary Relationship by Job Title")
```

------------------------------------------------------------------------

# Multiple Variable Analysis

Multiple variable analysis is a statistical technique that examines relationships between three or more variables simultaneously. The multiple variable analysis technique involves analyzing how multiple independent variables relate to one or more dependent variables, aiming to understand the interdependence of these variables and their impact on each other.

## Research questions

RQ1: Does gender, job title, and years of experience significantly impact an individual's annual salary?

RQ2: How much variance in salary can be explained by gender and job title after controlling for years of experience?

```{r, echo=F}
library(DiagrammeR)

DiagrammeR::grViz("
digraph relationships {
  graph [layout = dot, rankdir = LR]

  # Define nodes
  node [shape = ellipse, style = filled]
  X1 [label = 'gender (X1)', color = lightblue]
  X2 [label = 'job title (X2)', color = lightblue]
  X3 [label = 'experience (X3)', color = lightblue]
  Y [label = 'annual salary (Y)', fillcolor = lightblue, fontcolor = black]

  # Define edges (relationships)
  X1 -> Y
  X2 -> Y
  X3 -> Y
}
")

```

## Base statistical models

Below, I provide the base statistical models to answer the listed research questions. However, be reminded that you should utilize your theoretical framework to drive your analysis. The most important component of our modeling here are the post-hoc tests of our assumptions.

### Research Question 1

RQ1: Does gender, job title, and years of experience significantly impact an individual's annual salary?

```{r, message=F, warning=F}
# Fit the model
model <- lm(annual_salary ~ gender + job_title + years_experience, data = pay_gap_data)

# Get the summary of the model
summary(model)
```

#### Checking Assumptions

We then want to check the assumptions for our multiple linear regression model.

We use the `which` command here to call for specific results.

```{r, message=F, warning=F}
# Linearity and homoscedasticity
plot(model, which = 1)
```

```{r, message=F, warning=F}
# Normality of residuals
plot(model, which = 2)
```

```{r, message=F, warning=F}
# Influential points
plot(model, which = 4)
```

```{r, message=F, warning=F}
# Multicollinearity
vif(model)
```

```{r, message=F, warning=F}
# Heteroscedasticity test
bptest(model)
```

### Research Question 2

RQ2: How much variance in salary can be explained by gender and job title after controlling for years of experience?

```{r, message-F, warning=F}
# Extract R-squared values
r_squared_full <- model$r.squared

r_squared_experience <- 
  summary(lm(annual_salary ~ years_experience, 
             data = pay_gap_data))$r.squared

# Calculate the additional variance explained
variance.model <- r_squared_full - r_squared_experience
variance.model
```

### Additional tests to examine differences by groupings

```{r, message=F, warning=F}
## ANOVA: Job Title vs Salary
summary(aov(annual_salary ~ job_title, data = pay_gap_data))
```

```{r, message=F, warning=F}
## Kruskal-Wallis test (a non-parametric alternative)
kruskal.test(annual_salary ~ job_title, data = pay_gap_data)
```

```{r, message=F, warning=F}
## ANOVA for Gender Differences (3 groups)
# For normally distributed data
gender_salary_aov <- aov(annual_salary ~ gender, data = pay_gap_data)
summary(gender_salary_aov)

gender_experience_aov <- aov(years_experience ~ gender, data = pay_gap_data)
summary(gender_experience_aov)

## Post-hoc pairwise comparisons (if ANOVA is significant)
# Tukey's Honest Significant Differences
TukeyHSD(gender_salary_aov)
TukeyHSD(gender_experience_aov)

## Non-parametric alternative (Kruskal-Wallis)
kruskal.test(annual_salary ~ gender, data = pay_gap_data)
kruskal.test(years_experience ~ gender, data = pay_gap_data)

## Pairwise Wilcoxon Rank Sum tests with Bonferroni correction
pairwise.wilcox.test(pay_gap_data$annual_salary, pay_gap_data$gender,
                     p.adjust.method = "bonferroni")
pairwise.wilcox.test(pay_gap_data$years_experience, pay_gap_data$gender,
                     p.adjust.method = "bonferroni")

## Enhanced gender-salary plot
ggplot(pay_gap_data, aes(x = gender, y = annual_salary, fill = gender)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.4) +
  stat_compare_means(method = "anova", label.y = max(pay_gap_data$annual_salary)*1.1) +
  labs(title = "Salary Distribution Across Gender Groups",
       subtitle = "Three-group comparison with ANOVA results")

## Faceted experience-salary relationship
ggplot(pay_gap_data, aes(x = years_experience, y = annual_salary)) +
  geom_point(aes(color = gender)) +
  geom_smooth(method = "lm") +
  facet_wrap(~gender, ncol = 3) +
  labs(title = "Experience-Salary Relationship by Gender Group")
```

------------------------------------------------------------------------

# Multivariate Analysis

Multivariate analysis and multiple variable analysis are often confused, but they have a key distinction. Multivariate analysis specifically examines multiple dependent or outcome variables simultaneously, while multiple variable analysis (also called multivariable analysis) focuses on a single outcome variable with multiple independent variables.

## Research question

RQ1: Do gender and job title jointly influence both salary and years of experience?

RQ2: Are there significant differences in salary distributions across job titles when accounting for gender and experience?

```{r, echo=F}
grViz("
  digraph variables_diagram {
    graph [rankdir = LR]
    
    node [shape = ellipse, style = filled, fillcolor = lightblue]
    gender [label = 'Gender']
    job_title [label = 'Job Title']
    
    node [shape = ellipse, style = filled, fillcolor = lightyellow]
    salary [label = 'Annual Salary']
    experience [label = 'Years of Experience']
    
    gender -> salary
    gender -> experience
    job_title -> salary
    job_title -> experience
  }
")
```

### MANOVA

RQ1: Do gender and job title jointly influence both salary and years of experience?

To answer RQ1, we'll first run a multiple analysis of variance (MANOVA) on our data.

```{r, message=F, warning=F}
manova_result <- manova(cbind(annual_salary, years_experience) ~ gender + job_title, data = pay_gap_data)

summary(manova_result)
```

```{r, message=F, warning=F}
# Perform univariate ANOVAs for each dependent variable
summary.aov(manova_result)
```

```{r, message=F, warning=F}
# Post-hoc tests for job title (if significant)
TukeyHSD(aov(annual_salary ~ job_title, data = pay_gap_data))
TukeyHSD(aov(years_experience ~ job_title, data = pay_gap_data))
```

#### Check MANOVA assumptions

```{r, message=F, warning=F}
# Multivariate normality
mshapiro.test(t(model.matrix(manova_result)[, -1]))
```

```{r, message=F, warning=F}
# Homogeneity of covariance matrices
leveneTest(annual_salary ~ gender * job_title, data = pay_gap_data)
leveneTest(years_experience ~ gender * job_title, data = pay_gap_data)
```

### Visual of the results

```{r, message=F, warning=F}
ggplot(pay_gap_data, aes(x = job_title, y = annual_salary, fill = gender)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Salary Distribution by Job Title and Gender",
       x = "Job Title", y = "Annual Salary")

ggplot(pay_gap_data, aes(x = job_title, y = years_experience, fill = gender)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Years of Experience Distribution by Job Title and Gender",
       x = "Job Title", y = "Years of Experience")
```

RQ2: Are there significant differences in salary distributions across job titles when accounting for gender and experience?

To answer research question two, we'll run some additional analyses.

```{r, message=F, warning=F}
model_salary <- lm(annual_salary ~ job_title + gender + years_experience, data = pay_gap_data)
anova(model_salary)
summary(model_salary)
```

As noted in class, these are base models to help you consider your analysis. However, your theory should lead all of your work. Please let me know if you have any questions.
