---
title: "Univariate analysis"
author: "Nathan Alexander"
output: html_document
editor_options:
    chunk_output_type: console
---

```{r setup, include=F}
knitr:: opts_chunk$set(echo = TRUE)
```

# Libraries and packages

Load the libraries we needed. Be sure to `install.packages()` where needed.

```{r}
library(ggplot2)
library(dplyr)
```

### Task 1: Open a new RScript

Now we will open a new RScript.

Navigate to `File > New File > RScript`.

This RScript file is what you will use to outline your analysis.

### Task 2: Write a preamble

A preamble is similar to the heading of a paper.

The preamble contains information that will be useful for you and your collaborators.

```{r}
# Use the `#` symbol to tell R to ignore the text
## Name: <include your full name>
## Date: <sometimes you may want to add a date>
## Purpose: <insert the goals or purpose of the RScript>
```

Remember to use the `#` symbol to write your preamble.

------------------------------------------------------------------------

### Task 3: Check your working directory

Prior to inserting any code, it is generally helpful to check your working directory.

This will ensure that you are in the right location to call and save files.

```{r}
#| echo: true
#| output: false
# get the working directory
getwd()
```


------------------------------------------------------------------------

### Task 4: Install your packages

We will use the `install.packages` function to complete a few tasks.

Any required packages are generally placed at the very top of our code.

```{r}
#| echo: true
#| output: false
#| warning: false
install.packages("tidyverse", repos = "http://cran.us.r-project.org")
install.packages("remotes", repos = "http://cran.us.r-project.org")
```

------------------------------------------------------------------------

### Task 5: Load your libraries

After using the `install.packages` function, we need to load specific libraries.

Similar to installing packages, we generally load libraries at the very top of our code.

```{r}
#| echo: true
#| output: false
#| warning: false
# load libraries
library(tidyverse)
library(remotes)
```

------------------------------------------------------------------------

### Task 6: Call in the `critstats` data package

We will use the `remotes::install_github()` command to install packages to call in data that I have prepared in a package for our course. The `remotes::install_github()` command communicates with GitHub to access files.

```{r}
#| echo: true
#| output: false
#| warning: false
remotes::install_github("professornaite/critstats", force=TRUE)
```

------------------------------------------------------------------------

#### Task 6-a: Load the `critstats` library

Next, we load the `critstats` library which will give us access to the package's contents.

```{r}
#| echo: true
#| output: false
#| warning: false
library(critstats)
```

------------------------------------------------------------------------

### Task 7: View datasets in the current R session

RStudio has a lot of pre-loaded data sets that we can view or use to practice on different types of variables. We use the `data()` function to view all of the data sets loaded in R.

```{r}
#| echo: true
#| output: false
#| warning: false
data()
```

A new window should open in your RStudio session.

------------------------------------------------------------------------

### Task 8: View datasets only in the `critstats` data package

We can also view only the data sets loaded into the `critstats` package.

```{r}
#| echo: true
#| output: false
#| warning: false
data(package="critstats")
```

A new window containing only the `critstats` data should open in your RStudio session.

------------------------------------------------------------------------

### Task 9: View documentation for another data package called `datasets`

```{r}
#| echo: true
#| output: false
#| warning: false
data(package="datasets")
```

------------------------------------------------------------------------

### Task 10: View documentation for the entire `critstats` data package

```{r}
#| echo: true
#| output: false
#| warning: false
??critstats
```

------------------------------------------------------------------------

### Task 11: View documentation for a specific data set

We will select the `africa_data_all` data set and view its documentation.

```{r}
#| echo: true
#| output: false
#| warning: false
??africa_data_all
```

This documentation is the `codebook` for the data. It contains more specific information about the data frame, each of the variables, and any sourcing information.

------------------------------------------------------------------------

### Task 12: Assign a dataset to an object for efficiency

If we want to call a data set more efficiently, we can assign it to an object.

The `df` object is a short name for data frame.

We will assign `df1` using our assignment operator to the data set.

```{r}
#| echo: true
#| output: false
#| warning: false
df1 <- africa_data_all
```

------------------------------------------------------------------------

### Task 13: Run a simple command

We can run simple commands on a data frame using the shorthand object we assigned to it.

Let's use the `glimpse()` function to explore the `africa_data_all` data frame that we assigned to the object `df1`.

```{r}
#| echo: true
#| output: true
#| warning: false
glimpse(df1)
```

------------------------------------------------------------------------

### Task 14: View your data

We can use the `view()` function to see our entire data frame.

```{r}
#| echo: true
#| output: false
#| warning: false
view(df1)
```

You may have noticed that a new window opened so that you can `view()` the entire data set. Sometimes, it is inefficient to use this command. Let's look at two other options.

------------------------------------------------------------------------

#### Task 14-a: View the `head` of your data

We use the `head` function to get a closer look at the first few observations.

```{r}
#| echo: true
#| output: true
#| warning: false
# use the `head` function to view the top of the data
head(df1)
```

Change the number of observations that you want to view in the data frame.

```{r}
#| echo: true
#| output: true
#| warning: false
# use the `head` function to view ten (10) observations at the top of the data
head(df1, n=10)
```

------------------------------------------------------------------------

#### Task 14-b: View the `tail` of your data

We use the `tail` function to get a closer look at the last few observations.

```{r}
#| echo: true
#| output: true
#| warning: false
# use the `tail` function to view the top of the data
tail(df1)
```

Change the number of observations that you want to view in the data frame.

```{r}
#| echo: true
#| output: true
#| warning: false
# use the `tail` function to view ten (10) observations at the top of the data
tail(df1, n=10)
```
